---
title: "Elections and Longevity"
author: "David Kane"
format: html
execute: 
  echo: false
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(primer.data)
library(brms)
library(tidybayes)
library(gtsummary)

x <- governors |> 
  select(death_age, election_age, win_margin, party, sex) |> 
  mutate(treatment = ifelse(win_margin > 0, "win", "lose")) |> 
  filter(abs(win_margin) < 5) |> 
  drop_na()
```



```{r}
#| cache: true
#| warning: false

fit_elec_2 <- brm(formula = death_age ~ treatment + win_margin + 
                  party, 
                  data = x, 
                  family = gaussian(), 
                  refresh = 0, 
                  silent = 2, 
                  seed = 9) 
```

```{r}
# fixef(fit_elec_2)
```

Using data about US governor candidates from the years 1945 - 2012, we seek to find the relationship between the longevity of Preceptor David Kane and whether or not he wins the Mayoral Election in Newton, MA. Modern medicine has increased the overall lifespan of all candidates regardless of if they won or lost. We modeled age of death as a sum of election result (won/lost), age during election, political party and the win margin. We expect Preceptor to live an extra 8 years (plus/minus 6 years) if he were to win the Mayorial election.

$$death\_age_i = \beta_{0} + \beta_{1} treatment_i + \beta_{2}win\_margin_i + \\ \beta_{4}repub_i +  \beta_{5}thirdparty_i + \epsilon_{i}$$


```{r}
#| warning: false
fit_elec_2 |> 
  tbl_regression()
```

